Generate Real and Pseudo-absences

### 2023-12-19 scripts/generate_absences.Rmd
notes on 0 variance error trip

out <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)
out <- bind_rows(out)
st_write(out, '../data/raw/occurrence/absences.shp')

ab <- st_read('../data/raw/occurrence/absences.shp')

table(records$taxon)
```

@@

library(klaR)
library(psych)
library(MASS)
library(ggord)

# to determine whether the pseudo-random absences are in areas which are possibly suitable
# we will use linear discrimination analysis (LDA). LDA uses a label, e.g. presence or absence
# as a response which can be predicted by independent variables which are linear combinations 
# generated by reducing a higher dimensional data set to two axis. LDA will be trained on a data set 
# of AIM Plots with true absences, geographic absences, and a subset of the Psuedo-absences. 
# Only up to 750:750 presences:absences per species will be used to train a model. The maximum
# imbalance between presences to absences will be 750:250, or 3:1.

lda_PA_dropper <- function(x, path){

  pres <- x[x$Occurrence == 'Presence', ]
  real_abs <- x[x$PtType == 'AIM Absence', ]
  geo_abs <- x[x$PtType == 'Background Absence', ]
  model_abs <- dplyr::bind_rows(real_abs, geo_abs)

  if(nrow(pres) > 750){pres <- sample(1:nrow(pres), size = 750, replace = FALSE)}
  if(nrow(model_abs) > 750){model_abs <- sample(1:nrow(model_abs), size = 750, replace = FALSE)}

  training <- dplyr::bind_rows(pres, model_abs)
  testing <- filter(x, ! ID %in% training$ID)

  LDA_model <- MASS::lda(Species ~ ., training) # generate model

  # Here the accuracy of the trained modeled is evaluated
  LDA_train <- predict(LDA_model, training)$class
  tab <- table(Predicted = LDA_train, Actual = training$Species)
  ACC_train <- sum(diag(tab))/sum(tab) # accuracy

  # Here the model is applied to unclassified data
  LDA_test <- predict(LDA_model, testing)$class
  testing$Predicted <- predict(LDA_model, testing)$class
  tab1 <- table(Predicted = LDA_test, Actual = testing$Species)
  ACC_test <- sum(diag(tab1))/sum(tab1) # accuracy.

  n_pres <- nrow(x[x$Occurrence == 'Presence',])
  n_abs <- nrow(x[x$Occurrence != 'Presence',])
  conflicted <- filter(testing, Predicted != Occurrence) 
  max_rcs_2_remove <- n_pres - n_ab

  if(nrow(conflicted) == 0){ # identify the records we want to remove from the test data. 
    removals <- NA
  } else if (nrow(conflicted) > max_rcs_2_remove) {
    removals <- conflicted[sample(1:nrow(conflicted), size = max_rcs_2_remove, replace = FALSE), 'ID']
  } else {
    removals <- conflicted$ID
  }

  if(!is.na(removals)){data_out <- dplyr::filter(x, ! ID %in% removals)}

  metrics <- data.frame(
    Presences = nrow(pres), 
    Absences = nrow(model_abs),
    ACC_train = ACC_train, 
    ACC_test = ACC_test,
    Flagged = nrow(conflicted)
    )

  species <- x$Taxon[1]
  message(length(removals), ' records dropped from ', species)
  species <- gsub(' ', '_', species)
  write.csv(metrics, file.path(path, paste0(species, '.csv')), row.names = FALSE)

  return(data_out)
}

```

>>>>

out1 <- lapply(records, absence_drawer, terrestrial = land,  bg_abs = 0.15)

out <- bind_rows(out1) %>% 
  rename(taxon = Taxon) %>% 
  mutate(date = as.Date('2023-12-18'), .before = geometry)

st_write(out, '../data/raw/occurrence/absences/absences.shp', append = FALSE)

rm(out1, blm_surf, AIM_points, records)
```

```{r import raster stack}

p <- '/media/steppe/hdd/Geospatial_data/WesternPlantPredictors2/chelsa'
paths <- file.path(p, list.files(p, recursive = F))
files <- lapply(paths, function(x){ file.path(x, list.files(x)) })
names(files) <- basename(paths)
bio <- lapply(files, vrt)
bio <- rast(bio)
bio <- subset(bio, c(1:11, 14:17, 19)) # these 'empty' need investigate

rm(p, paths, files)
```

```{r subset rasterstack, eval = F}
vif_vals <- spatSample(bio, size = 5000, values = TRUE) # this steps takes forever... 
vif_vals <- vif_vals[ complete.cases(vif_vals), ] # 4148 records used
collinear_probs <- usdm::vifstep(vif_vals, th = 10)
bio <- usdm::exclude(bio, collinear_probs)

write_rds(collinear_probs, '../results/bioclim_VIF_PA_drop.rda')

rm(p, paths, files)
```

```{r import potential records}

absences <- st_read('../data/raw/occurrence/absences/absences.shp', quiet = TRUE)
presences <- st_read('../data/raw/occurrence/thinned/thinned_occurrences.shp', quiet = TRUE) %>% 
  mutate(Occurrence = 1, 
         PtType = if_else(is.na(PrmryKy), 'AIM Presence', 'Opportunistic Presence'), .before = geometry)

recs <- bind_rows(absences, presences) %>% 
  group_by(taxon) %>% 
  mutate(ID = 1:n())

collinear_probs <- read_rds( '../results/bioclim_VIF_PA_drop.rda')
bio <- usdm::exclude(bio, collinear_probs)

rm(absences, presences, collinear_probs)
```

@@

set.seed(27)

cn <- names(bio)
# splicies <- split(recs, f = recs$taxon)
lapply(splicies, lda_PA_dropper, path = '../results/lda/', col_names = cn)

which(names(splicies)[130] == 'Dudleya cymosa')
# need to restart on record 130, 
# which is Dudleya cymosa ssp. cymosa, and which should be subsumed into Dudleya cymosa!
# the problem was that a variable displayed 0 variance because there were only 2 records!!!
```
